---
title: Using Python and R
author: ''
date: '2020-05-09'
slug: using-python-and-r
categories: []
tags: []
description: ''
---
One can easily switch between Python and R studio through the reticulate package. The reticulate package allows the R interface to run Python code.

# Setup
Open up an R-markdown file to explore. First, install the reticulate package to and {R} code chunk by running the following code install.package() code and then calling the package through library().
```{r}
library(reticulate)
```

# Basic Example
The next step is to assign information into a object.  Here I assigned 'dog' into 'd'. 
```{r}
#R Studio
d<-"dog"
```

Next step is to create a {python} code chunk. Now assign information, similarly to the R code. For this example I assigned 'cute' into 'd'. Then to allow for the exhange of information between the chunks, use the function 'print()'. Using this function you can access the previously assigned R objects with 'r.objectname'. The 'print()' function looks for things in the R environment and reads then in python. Above I used 'd' for the R code, so to access it within 'print()' I just added 'r.d'. The R and python code interact with each other, as seen by the output. 
```{python}
#Python
d='cute'
print(d,r.d)
```

# Potential Uses 
A ootential use is pulling an R package and using ut within python. This can allow you to import an dataset that is built-into R and use it in Python. Similarly, if there is a dataset or something of interest in R, you can pull it from python and use it in R.  